---
output:
  html_document: default
  pdf_document: default
---
Red Wine by Nadim Kawwa
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load the libraries

library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(knitr)
library(GGally)
library(MASS)
library(lattice)
library(latticeExtra)
library(colorspace)
library(RColorBrewer)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityReds.csv')
```

```{r echo=FALSE, explore}
summary(wine)

```

> Take a quick look at the data. At a glance we can tell that most wines are
acidic, and the alcohol mean is more or less equal to the median.
The section below will plot these variables individually and interpret the
output.


# Univariate Plots Section

>Start by plotting parameters in order to see their distribution.

```{r echo=FALSE, quality}

ggplot(data = wine, aes(quality)) +
  geom_bar()+
  xlab('Quality')+
  ylab('Count')+
  ggtitle('Distribution of Red Wine Quality')+
  geom_vline(aes(xintercept=mean(wine$quality)), color='red')


```

>The quality is between 3 and 8, with most of the winesrated a 5 or a 6. It hence makes sense that the average quality is somewhere in that range as seen with the red line.


```{r echo=FALSE, pH}

ggplot(data = wine, aes(wine$pH)) +
  geom_histogram(binwidth = 0.1)+
  xlab('pH')+
  ylab('Count')+
  ggtitle('Distribution of Red Wine pH')


```

> The pH looks normally distributed. For this first part we're not elemenating
any outliers. Note that there's an outlier with a pH close to 1 (is that wine
or water?)

```{r echo=FALSE, alcohol}

ggplot(data = wine, aes(alcohol)) +
  geom_histogram(binwidth = 0.5)+
  scale_x_continuous(limits = c(9,14))+
  xlab('Alcohol Content (% by volume)')+
  ylab('Count')+
  ggtitle('Distribution of Red Wine Alcohol Content')
```


> Alcohol content is clearly skewed to the right, with higher alcohol contents
the rarer of the bunch. Note that for alcohol content we don't know how large
the wine bottle (or keg?) is.

```{r echo=FALSE, acidity}

ggplot(data = wine, aes(wine$fixed.acidity)) +
  #Range is 4-16, fitting in 0.5 bins should give a better visual.
  geom_histogram(binwidth = 0.5)+
  xlab('fixed acidity (tartaric acid - g / dm^3)')+
  ylab('Count')+
  ggtitle('Distribution of Red Wine Fixed Acidity')
```

> The fixed acitidity looks like it could be skewed to the left. We also notice
how it has a single peak. For a better visualisation let's go to the next plot.


```{r echo=FALSE, acidbox}
ggplot(data=wine, aes(x=1,y=fixed.acidity))+
  geom_jitter(alpha = 0.1, colour='black')+
  #Fit a box plot over the nodes to see fixed acidity distributoin.
  geom_boxplot(color='darkblue', alpha = 0.5)+
  ylab('fixed acidity (tartaric acid - g / dm^3)')+
  xlab('')+
  ggtitle('Fixed Acidity Distribution')

```

> To take another look at fixed acidity draw a box plot. On top of that overlay 
the scattered points of tbe acidity values. Most acidity values are between 7
and 9.


```{r echo=FALSE, density}

ggplot(data = wine, aes(wine$density)) +
  #Density  ranges from 0.990-1.000. Calls for very small bins
  geom_histogram(binwidth = 0.0005)+
  scale_x_continuous(limits = c(0.990, 1.005))+
  xlab('density (g / cm^3)')+
  ylab('Count')+
  ggtitle('Distribution of Red Wine Density')
```

> Density is another metric to measure. What the plot tells is that the density 
is normally distributed, and that most wines have a density more or less equal 
to water.

```{r echo=FALSE, citric, warning=FALSE, message=FALSE}
ggplot(data=wine, aes(x=citric.acid))+
  geom_histogram(binwidth = 0.05, colour="black", fill="grey")+
  scale_x_continuous(limits = c(-0.05, 0.8))+
  xlab("citric acid (g / dm^3)")+
  ylab("Count")+
  ggtitle("Distribution of Citric Acid")

```

> Citric acid is what gives wine its sour taste. We don't see much of a trend
in the distribution.

```{r echo=FALSE, sugar, warning=FALSE, message=FALSE}
ggplot(data=wine, aes(x=residual.sugar), binwidth=0.25)+
  geom_histogram(colour="black", fill="orange", binwidth = 0.2)+
  scale_x_continuous(limits = c(2,6))+
  xlab("residual sugar (g / dm^3)")+
  ylab("Count")+
  ggtitle("Red Wine Residual Sugar")

```

> Residual sugar content is clearly skewed to the right with a peak at 2.2.
Are most wines on the lower sugar content side?

```{r echo=FALSE, chlorides, warning=FALSE, message=FALSE}
ggplot(data=wine, aes(x=chlorides))+
  #Same as density, use small bins
  geom_histogram(colour="black", fill="red", binwidth = 0.005)+
  #Long and flat tail after 0.2, set it as the limit
  scale_x_continuous(limits = c(0, 0.2))+
  xlab("chlorides (sodium chloride - g / dm^3")+
  ylab("Count")+
  ggtitle("Red Wine Chlorides")

```

> Clorides are what give the wine it's salty taste. The distribution is skewed
to the right with most wines having very little salt content. The long tail
is between 0.2-0.6 and was omitted for this plot.

```{r echo=FALSE, volatileacid}
ggplot(data=wine, aes(x=volatile.acidity))+
  geom_histogram(colour="black", fill="yellow", binwidth = 0.02)+
  xlab("volatile acidity (acetic acid - g / dm^3)")+
  ylab("Count")+
  ggtitle("Red Wine Volatice Acidity")

```

> Volatile acidity is what gives the wine its vinegar taste. We can tell from 
the plot that the distribution is bimodal, with one peak at 0.4 and another at
0.6.


```{r echo=FALSE, SO2}
ggplot(data=wine, aes(x=total.sulfur.dioxide))+
  geom_histogram(colour="black", fill="brown", binwidth = 10)+
  xlab("total sulfur dioxide (mg / dm^3)")+
  ylab("Count")+
  ggtitle("Red Wine Total Sulfure Dioxide")+
  geom_vline(aes(xintercept=median(wine$total.sulfur.dioxide)), color='blue')
```

> Above 50ppm SO2 becomes detectable by smell. for water 1ppm = 1mg/L with
1 liter = 1 dm^3. Let's assume wine is like water in this respect since their densities are almost equal. The distribution is skewed to the right, less than
half of the wines can be tested for SO2 by smell. 


# Univariate Analysis


> The dataset contains a collection of parameters measured off red wine 
samples.
The advantage of these observations is that they are measured in discrete 
values that can be measured. For example saltiness is measured in SO2 
concentration rather than subjective terms (salty, not very salty, etc...)
We will try to see if there is a correlation between these features and the
quality of the wine.
It will be of interest to see if there is a correlation among these paramters
themselves.
Among the paramters only citric acid content did not show a distinct pattern.
Some of the plots with longer tails were trimmed in order to better show
the distribution near the peak.
At this point no new variables were created.


# Bivariate Plots Section


```{r echo=FALSE, cor}

#Let's take alcohol, citric acid, sugar, quality, and quality and put them in a 
#small and more manageable dataframe

wine_small <- wine[, c("alcohol","citric.acid", "residual.sugar", "pH", 
                       "quality")]

ggpairs(wine_small)


```

> We took four arbitrary parameters that should influence quality. The strongest
correlation for quality exists for alcohol content. pH seems to have almost no
effect on quality which sounds counterintuitive. For all wines the quality of
the soil definetly influences the quality of the grapes.
pH and citric acid have a negative correlation which makes sense. The more 
acidic the solution the lower the pH.



```{r echo=FALSE, alcohol_vs_quality}
ggplot(data = wine, aes(x=alcohol, y=quality))+
  geom_jitter(alpha = 0.25) +
  xlab("Alcohol by Volume (%)")+
  ylab("Quality")+
  ggtitle("Alcohol by Volume Vs Quality")+
  coord_cartesian(xlim = c(quantile(wine$alcohol, 0.05),
                            quantile(wine$alcohol, 0.95))) +
  geom_smooth(method = "lm")

```

> Try plotting alcohol vs quality and fit in a linear model regression. 
As expected the slope is positive. This does excludes the top 5% and bottom 5%
of alcohol content. 

```{r echo=FALSE, alcohol_vs_quality_box}
ggplot(data = wine, aes(x=factor(quality), y=alcohol)) +
  geom_jitter(alpha = 0.1) +
  geom_boxplot(alpha= 0.1, color = "darkblue")+
  stat_summary(fun.y="mean", geom="point", color= "red")+
  xlab("Quality")+
  ylab("Alcohol by Volume (%)")+
  ggtitle("Quality Vs Alcohol Content")

```

> This box plot representaion of the same data offers a better insight.
We can guess that most wines have a quality between 5 and 6 as indicated by the
contration of data points. More noticeable is that the higher quality wines 
(score 7-8) have a higher alcohol content. However we can't ignore one 
pertinent outlier: Wines rated as 5 have a lower mean alcohol contnet than those 
rated 3-4. It would have been convenient to draw a neat line connecting the
means however that doesn't seem to be plausible.


```{r echo=FALSE, fixed_vs_volatile}
ggplot(data = wine, aes(x=fixed.acidity, y=volatile.acidity))+
  geom_jitter(alpha = 0.25) +
  xlab("Fixed Acidity")+
  ylab("Volatile Acidity")+
  ggtitle("Fixed vs Volatile Acidity")+
  coord_cartesian(ylim = c(quantile(wine$volatile.acidity, 0.05),
                           quantile(wine$volatile.acidity, 0.95)))+
  geom_smooth(method = "lm")

```

> Contrary to the names, these two are not inversly related. Take out the top 
and bottom 5% and what you get is the plot above.


```{r echo=FALSE, quality_vs_sugar}
ggplot(data=wine, aes(x=residual.sugar, y=quality)) +
  geom_jitter( alpha = .3)+
  coord_cartesian(xlim = c(0, quantile(wine$residual.sugar, 0.9)))+
  xlab(" Residual Sugar")+
  ylab("Quality")+
  ggtitle("Quality Vs Residual Sugar")+
  geom_smooth(method = "lm")

```

> Clearly there is no correlation here, the nodes are arranged in a homogenous fashion. Adding a linear model regression shows a horizontal line which again
confirms our observation. One shortcoming of this is the omission of dessert
wine by omitting the top 10% sweetest wines.
We will plot again for the sweetest wines.


```{r echo=FALSE, quality_vs_sugar_part2}
ggplot(data=wine, aes(x=residual.sugar, y=quality)) +
  geom_jitter( alpha = .3)+
  coord_cartesian(xlim = c(quantile(wine$residual.sugar, 0.9), quantile(wine$residual.sugar, 1)))+
  xlab(" Residual Sugar")+
  ylab("Quality")+
  ggtitle("Quality Vs Residual Sugar for Sweet Wines")+
  geom_smooth(method = "lm")

```

> Not much of a difference here. Turns out the sweetness of the wine is
irrelevant for us.

```{r echo=FALSE, quality_vs_SO2, warning=FALSE, message=FALSE}
ggplot(data=wine, aes(x=total.sulfur.dioxide, y=quality)) +
  geom_jitter( alpha = .3)+
  scale_x_continuous(limits = c(0, 150))+
  xlab(" Total Sulfure Dioxide")+
  ylab("Quality")+
  ggtitle("Quality Vs Total Sulfure Dioxide")+
  geom_smooth(formula = 'y ~ poly(x,2')

```

> By omotting the more pungent wines (SO2 > 150ppm), we can see that quality
and total sulfure dioxide are negatively correlated. This is using a second
order polynomial fit although the relationship seems linear. 
SO2 has a suffocating smell and higher concentrations negatively impact the
quality of the wine.

> Not much of a difference here. Turns out the sweetness of the wine is
irrelevant for us.

```{r echo=FALSE, quality_vs_volatile}
ggplot(data=wine, aes(x= volatile.acidity, y=quality)) +
  geom_jitter( alpha = .3)+
  scale_x_sqrt()+
  xlab("Sqrt(Volatile Acidity)")+
  ylab("Quality")+
  ggtitle("Quality Vs Volatile Acidity")+
  geom_smooth(method = "lm")

```

> There seems to be a negative correlation between volatile acidity and wine
quality. Since volatile acidity is mostly smaller than 1, apply a square
root transform to obtain larger values and maybe spread out the scatter.



# Bivariate Analysis


>At this point it is worth moving into multivariate analysis. We will explore
alcohol content, volatile acidity, and total sulfure dioxide. These four appear
to be the strongest in terms of correlation.The strongest relationship for determining the quality of a wine appears to be alcohol content.
What is missing from the above plots is one or more variables with a 
correlation coefficient larger than 0.7 to be considered significant.

# Multivariate Plots Section


```{r echo=FALSE, quality_alcohol_SO2}

ggplot(data = wine, aes(x=alcohol, y=quality, color= total.sulfur.dioxide)) +
  #Jitter the points to avoid overcrowding
  geom_point(position = "jitter",alpha = 0.5, size =2)+
  scale_color_gradient(high = 'black', low = 'grey')+
  xlab("Alcohol Content")+
  ylab("Quality")+
  ggtitle("Alcohol vs Quality Colored by SO2 Content")

```

> Not much of a discernable color pattern here. This suggests that SO2 concentration isn't signficant enough if coupled with alcohol content.


```{r echo=FALSE, quality_alcohol_citric}

ggplot(data = wine, aes(x=alcohol, y=factor(quality), color= citric.acid)) +
  #Jitter the points to avoid overcrowding
  geom_jitter(position = "jitter", alpha = 0.5, size =2)+
  scale_color_gradient(high = 'black', low = 'grey')+
  xlab("Alcohol Content")+
  ylab("Quality")+
  ggtitle("Alcohol Vs Quality Colored by Citric Acid")
```

> The plot above colored by citric acid content shows more. The higher quality
wines are more brightly colored which suggests that citric acid couipled
with alcohol content makes for a higher quality wine.

```{r echo=FALSE, quality_alcohol_ph, warning=FALSE, message=FALSE}

ggplot(data = wine, aes(x=alcohol, y=pH, color = factor(quality))) +
  geom_point(alpha = 0.5, size =2) +
  xlab("Alcohol Content")+
  ylab("pH")+
  ggtitle("Alcohol vs pH Colored by Quality")+
  scale_fill_brewer()+
  geom_smooth()+
  theme_dark()+
  scale_color_discrete(name = "Quality")

```

> There seems to be a positive correlation between alcohol and pH value. That
being said most wines have a pH between 3 and 3.5. 
If we look at the  linear fit for wines between 7 to 8 quality rating they are 
on the lower side of pH value. Conversly wines with a poor 3-4 quality are in
the upper part of the plot signaling a higher pH value.
The oddities are at the extremeties of the fitten models. For qualitites 3 to 6
the linear models are overlapping at low alcohol content (less than 10%).
For heavier contents the qualities 6-8 seem to converge.



# Multivariate Analysis


>There seem to be two major factors that influence quality based on the above: alcohol content and citric acid concentration. The parameters are not
sufficiently impactful to influence the quality of the red wine. There might
be other factors at play such as the taste of the people giving the rating.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

ggpairs(wine_small)
```

### Description One
>The table above shows a correlation between arbitrary pairs. We can tell at
a glance that alcohol and citric acid will play a more significant role in
predicting the quality of the wine.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data = wine, aes(x=factor(quality), y=alcohol)) +
  geom_jitter(alpha = 0.1) +
  geom_boxplot(alpha= 0.1, color = "darkblue")+
  stat_summary(fun.y="mean", geom="point", color= "red")+
  xlab("Quality")+
  ylab("Alcohol by Volume (%)")+
  ggtitle("Quality Vs Alcohol Content")

```

### Description Two
> The alcohol content is plotted against quality in the form of a box plot.
This takes into a clearer perspective the 0.476 correlation coefficient sceen in
Plot_one. We can infer that a higher alcohol content means a better quality.
However we can't ingnore the dip in alcohol content seen at wines with a score
of 5. Does alcohol within a certain range interect badly with other elements
that give wine its taste? 


### Plot Three
```{r echo=FALSE, Plot_Three, warning=FALSE, message=FALSE}
ggplot(data = wine, aes(x=alcohol, y=pH, color = factor(quality))) +
  geom_point(position= "jitter", alpha = 0.5, size =2) +
  xlab("Alcohol Content")+
  ylab("pH")+
  ggtitle("Alcohol vs pH Colored by Quality")+
  scale_fill_brewer()+
  geom_smooth()+
  theme_dark()+
  scale_color_discrete(name = "Quality")

```

### Description Three
> The third plot which I thought would be interesting is pH vs Alcohol colored
by quality. First to notice are oddities. Wines ranging from 3 to 6 quality are
given the same linear fit; those with qualities 6 to 8 appear to converge
towards the end. 
Looking away from the edges the higher quality wines have a
low pH and the poorer quality wiens have a high pH. This observation is based
on the linear models fitted to the plot and may not be entirely accurate.
We don't believe that a large enough number of outliers to skew the fit.


------

# Reflection

> The dataset contains a large sample of wines with parameters that can be
objectively measured. Throughout the study an attempt was made to find if there
is a strong correlation between quality and one of the aspects
Surprisingly none of the parameters had a correlation factor bigger than 0.7, 
the arbitrary value beyond which a strong correlation is evident.
There are some limitations to the dataset, beginning by the identity of the
graders. A wine expert is supposedly a sommelier and there are levels to that
qualifications, does one understand more about quality than the others?
There is also the limitation about geographic origin. However what would be
more insightful is to take a single wine producer and track the quality of
their vintages over time.
